\chapter*{\runtitle}

\noindent
The \emph{Community Membership Hiding} (CMH) problem asks: given a graph, a community
detection algorithm, and a target node $u$, find a minimal set of edge rewirings such
that $u$ is no longer detected as a member of its original community. This problem arises
naturally in the context of network privacy, where community detection algorithms can
inadvertently expose sensitive user affiliations---such as political or ideological
leanings---solely from the structural position of a node in the underlying graph.

Recent work has formulated CMH as a Markov Decision Process (MDP) solved via Deep
Reinforcement Learning (DRL), employing an Actor-Critic architecture. While that
approach demonstrated the viability of the paradigm, we identify a critical architectural
limitation: the agent evaluates the network globally without explicitly conditioning on
the identity of the target node, thereby learning a generic community-shuffling policy
rather than a targeted hiding strategy.

In this thesis, we propose a novel DRL architecture that addresses these fundamental
shortcomings. Our model explicitly incorporates the target node as a feature of the
network representation, employs Graph Attention Networks v2 (GATv2) as the graph
encoder to leverage dynamic, query-conditioned attention, and combines static and
real-time structural signals to accurately track topological changes during rewiring.
We further extend the single-agent system into a \emph{Dual-Agent Architecture} that
dedicates separate agents to edge addition and edge deletion, and introduce a
\emph{Policy-Guided Subgraphing} heuristic that makes the framework tractable on
large-scale graphs.

Extensive experimental evaluation on the standard benchmark datasets of the field
demonstrates that our architecture consistently outperforms both the original DRL
agent and strong gradient-based baselines, achieving high hiding success rates with
minimal structural perturbation.

\bigskip

\noindent\textbf{Keywords:} community detection, community membership hiding, deep
reinforcement learning, graph attention networks, network privacy, GATv2, Actor-Critic.
